### 2012-03-09 draft/draft.tex
The introduction is added.

\documentclass[times, 10pt, twocolumn]{article}
\usepackage{latex8}
\usepackage[dvips]{graphicx}
\usepackage[fleqn]{amsmath}
\usepackage{amsthm}
\usepackage{txfonts}
\usepackage{courier}
\usepackage{subfigure}
\usepackage{comment}
\usepackage{url}
\usepackage[square,numbers]{natbib}

%-----------------------------------------
% need for camera-ready
\pagestyle{empty}

%----------------------------------------

\begin{document}

\title{
Toward Adaptive GPU Resource Management for Embedded Real-Time Systems
}

\author {
Junsung Kim and Ragunathan (Raj) Rajkumar\\
\\
Department of Electrical and Computer Engineering\\
Carnegie Mellon University
\and
Shinpei Kato and Scott Brandt\\
\\
Department of Computer Science\\
University of California, Santa Cruz
}

\maketitle

%-----------------------------------------
% need for camera-ready
\thispagestyle{empty}

\begin{abstract}
 In this paper, we present conceptual frameworks for GPU applications to
 adjust their task execution times upon given workloads.
\end{abstract}

\section{Introduction}
\label{sec:introduction}
Graphics processing units (GPUs) are widely used in a range of applications from supercomputers to mobile computers. For example, three of the top five supercomputers on the TOP500 list~\cite{TOP500} utilize GPUs to accelerate computations, and recent tablets such as ASUS Eee Pad Transformer Prime leverage 12 GPU cores of NVIDIA’s well-known SoC, Tegra 3 to boost the performance while saving power~\cite{Tegra3}. This trend will only continue and in fact is expected to accelerate across multiple application domains.

Modern automobiles have several tens of processing units, and advances in safe-driving features such as adaptive cruise control, stop-and-go cruise control, lane keeping and assisted lane change require even higher performance computing. As vehicles become semi-autonomous, drive autonomously on demand and eventually become fully autonomous, a multitude of computer vision, sensor fusion, signal processing and graphics problems must be solved in real time. Since solutions to those problems can leverage benefits of parallel processing techniques, GPUs can be used to accelerate computations. As an initial step, NVIDIA GPUs will be used for infotainment systems in future BMW vehicles~\cite{NVIDIA_BMW}.

Recent trends in self-driving cars~\cite{Urmson08, Markoff10, Kelly12} are motivated by various benefits of automated driving, but most importantly enabling automatic safety features. Those features are realized by the smart vehicle planning and the intelligent processing of data from various types of sensors on self-driving cars such as LIDAR (LIght Detection And Ranging), radar, camera, ultrasonic sensor, and so forth. One common characteristic of both the smart vehicle planning and the intelligent data processing is that their performances can be greatly accelerated using GPU, hence parallelism. For example, the more a self-driving car plans its potential paths to follow {\it a priori}, the better the quality of self-driving is. Since the self-driving car should follow only the best path, calculation of each potential path can happen parallel. A CMU team has already started using GPUs for autonomous driving computations. They have increased the performance of motion planning for autonomous driving up to 40 times using an NVIDIA GPU GTX 260 having 192 logical cores~\cite{McNaughton11}.

The perception system which intelligently processes the data from various sensors can also obtain benefits using GPU. The perception system of a self-driving car should be able to detect, classify and track the obstacles around itself. Various types of sensors will generate huge amount of information to understand the surroundings. In order to process the data, a parallel processing technique is of vital importance. For example, a new self-driving car at CMU manages 1536 objects from LIDAR sensors before they are fused with other types of sensor data. There has been on-going research using GPU to build a perception system~\cite{Ferreira11}, and their GPU implementation yielded 30,000-times-faster performance compared to the case of using only CPU.

As described above, there has been research on applying GPU to different applications on self-driving cars, and it is clear that GPU can provide great benefits on realizing self-driving car technologies. However, not much research has been done when those technologies are deployed together on self-driving cars, where the loads of each application dynamically vary depending on environments. The period and the computation time of the planning algorithms for autonomous driving highly depend on the vehicle speed, so the planning algorithms can be heavily loaded when the car is on highway. The load of the perception algorithms mainly depend on the number of obstacles around the car hence the perception system requires more computing resources when the car is driving in the urban area. Therefore, an intelligent method of sharing many cores on GPU would be essential when we use GPU on self-driving cars. For example, if a self-driving car has a 96-core GPU, the planning algorithm of the car can use 72 cores on the highway and use 12 cores in the urban environment. A self-driving car~\cite{Urmson08} require tens of tasks, and the dynamic core management should be fulfilled across all tasks if those tasks utilize GPU.

In this paper, we propose conceptual frameworks for GPU applications to adjust their task execution times depending on their given workloads.

The rest of paper is organized as follows. Section~\ref{sec:system_model} describes how our proposed system is modeled. Section~\ref{sec:adaptivity_support} presents the methods for adaptively managing the given GPU resources, and we conclude our paper in Section~\ref{sec:summary}.

\section{System Model}
\label{sec:system_model}

We assume such embedded systems that equip GPUs as compute devices.
The application task starts execution on the CPU, and offloads its
data-parallel compute-intensive workload onto the GPU when needed.
Once offloaded onto the GPU, the task becomes non-preemptive due to many
reasons.
In fact, it is technically possible to preempt the running task on the
GPU by loading and restoring its context, but it requires additional
firmware, runtime, and OS support, and the preemption cost would be
non-trivial due to a very large set of GPU registers and states.
We hence restrict our attention to a non-preemptive execution model for
GPU computing.
GPUs may also pose some constraints in multi-tasking.
Even the NVIDIA Fermi architecture~\cite{Fermi}, one of the most popular
GPU product lines, allows only one context to use GPU resources at
once, though this context may spawn multiple GPU kernels (jobs)
simultaneously.
In other words, if task-level parallelism is required, the entire system
must run in the same context.
We however believe that this constraint will not limit the concept of
this paper.
We can use the same context to exploit concurrent parallel job
executions, and it should be enough to demonstrate a proof of concept. 
We also expect that future product lines will remove this concern.

We consider real-time applications where each task runs in a periodic or
sporadic manner under deadline constraints.
Such a task set and application includes motion planning and 
vision-based perception in state-of-the-art autonomous driving
vehicles, where the periods often correspond to frame-rates, and the
deadlines appear at the end of period.
It should however be noted that deadline constraints may not be relevant
to periods depending on applications, but the computation must meet the
deadline in order to achieve the desired quality of service in either case.
We also presume that the computing demand of each task is highly
variable.
For example, the performance of planning and perception is usually
governed by the number of objects, the size of data, and the desired
quality of output.
These workloads are also well parallelizable using the GPU.
Hence, the autonomous driving tasks and applications are our primary
target, but the contribution of this paper is not limited to those but
is also generally applicable to highly vaiable workloads running on the
GPU.

\section{Adaptive GPU Resource Management}
\label{sec:adaptivity_support}

In this section, we provide an idea of adaptivity support for GPU
applications.
We particularly focus on computing resource allocation problems.
It is true that some embedded real-time systems exhibit highly variable
workloads.
Since GPUs integrate a great number of compute cores on a chip, we can
support even highly variable workloads in a timely manner by adjusting
allocated cores at runtime.

There are several levels of approaches for adaptive GPU resource
management.
Previous work~\cite{Kato_RTAS11, Kato_RTSS11, Kato_ATC11} took
time-driven approaches that control timings and the amounts of time
allowed to access GPU resources, \textit{i.e.}, scheduling and
reservation.
In these time-driven approaches, application tasks need not to be aware
of what is happening in GPU resource management, because it is handled
by the OS or runtime scheduler.
However, they can never manage task execution times.
They also limit the number of contexts that can access the GPU
simultaneously to remove performance interference.
Therefore, GPU resources could be wasted if a running context does not
fully use compute cores.

We consider a different approach than previous work that enables GPU
applications to adjust their task execution times upon workload.
We believe that the number of cores used in the program is a major
factor that could change the execution time, and hence explore how to
adjust it at runtime.
It is important to note that programmer is typically responsible for
allocating the number of cores (or threads mapped to cores) in GPU
programming.
In order to adjust the number of cores at runtime, it is essential to
provide the programmer with an interface to obtain the information on
the number of cores available or allocated for the program at runtime.
The programmer is then responsible for making the program adaptive to
the number of cores.

In the following, we present two frameworks that could achieve our idea.
Although they are still at an idea stage, we plan to implement a real
system as a proof-of-concept, leveraging open-source
software~\cite{Kato_OSPERT11}.

\subsection{Explicit Adjustment}

We propose an explicit adjustment framework where the programmer is
responsible for adjusting the number of cores to relax or tighten
the computing demand.
There will be no adjustment unless the programmer explicitly takes an
action.
A typical usage of this framework with periodic real-time tasks is as
follows:

At the end of each period, the programmer calls a API function provided
by our framework that returns the latest task execution time.
The programmer next calls either of the following two API functions.
One increases the number of cores to be used in the next period or GPU
invocation to speed up the program.
The other decreases it to speed down the program.
This framework would work fine because the programmer often knows the
desired task execution time to meet the frame-rate or deadline.
It is also flexible in that the programmer can determine when to
increase or decrease the number of cores.

A downside of this framework is that some task may misbehave
and interfer with other contending application tasks, if the programmer
fails to call the API functions correctly.
We can cap the maximum number of cores available for an individual task
to prevent it from abusing GPU resources, but adaptivity of computing
depends on programmers, and the system cannot really control it.

\subsection{Implicit Adjustment}

Our second approach to adaptive resource management is an implicit
adjustment framework.
In this framework, the number of cores to be allocated for the program
is set by the runtime system.
Hence, adaptivity of computing does not really depend on programmers,
but if the program is not aware of this framework, it may fail to run,
since the number of core allocated for the program may be different from
what the program assumes.

The programmer specifies the desired task execution time as a set point
before the task starts.
If this set point is not specified, the runtime system tries to derive
it internally as time goes by.
When the task uses the GPU, the runtime system consistently updates the
number of cores available for the corresponding task in the next period
based on the previous execution time records.
It is a programmer's duty to check the number of available cores before
offloading the computation onto the GPU.

This implicit adjustment framework is more encouraged than the explicit
adjustment framework, as it can enforce adaptive GPU resource
management.
However, it requires consensus in the programming model that the number
of cores allocated for the program could be changed every time it is
offloaded onto the GPU, and the programmer must be aware of it to make
the program work.
We claim that this is a natural trade-off between generality of
programming and adaptivity of service.

\section{Summary}
\label{sec:summary}

In this paper, we have discussed adaptivity requirements in embedded
real-time systems with GPUs, and presented two frameworks for adaptive
GPU resource management.
We conjecture that generality of programming may need to be compromised
to achieve adaptivity of resource allocation on the GPU.
Nonetheless, adaptive resource management is a key solution in
optimizing performance under resource-constrained environments.
We believe that our frameworks for GPU programming are useful
contributions in this line of work, given that trends are apparently
shifting to highly parallel heterogeneous computing.

\bibliographystyle{plain}
{\footnotesize
\bibliography{references}
}

\end{document}

### 2012-03-09 draft/draft.tex
first commit

@@
 \label{sec:introduction}
-Graphics processing units (GPUs) are widely used in a range of applications from supercomputers to mobile computers.
-For example, three of the top five supercomputers on the TOP500 list~\cite{TOP500} utilize GPUs to accelerate computations, and recent tablets such as ASUS Eee Pad Transformer Prime leverage 12 GPU cores of NVIDIA’s well-known SoC, Tegra 3 to boost the performance while saving power~\cite{Tegra3}.
-This trend will only continue and in fact is expected to accelerate across multiple application domains.
-
-Modern automobiles have several tens of processing units, and advances in safe-driving features such as adaptive cruise control, stop-and-go cruise control, lane keeping and assisted lane change require even higher performance computing.
-As vehicles become semi-autonomous, drive autonomously on demand and eventually become fully autonomous, a multitude of computer vision, sensor fusion, signal processing and graphics problems must be solved in real time.
-Since solutions to those problems can leverage benefits of parallel processing techniques, GPUs can be used to accelerate computations.
-As an initial step, NVIDIA GPUs will be used for infotainment systems in future BMW vehicles~\cite{NVIDIA_BMW}.
-
-Recent trends in self-driving cars~\cite{Urmson08, Markoff10, Kelly12} are motivated by various benefits of automated driving, but most importantly enabling automatic safety features.
-Those features are realized by the smart vehicle planning and the intelligent processing of data from various types of sensors on self-driving cars such as LIDAR (LIght Detection And Ranging), radar, camera, ultrasonic sensor, and so forth.
-One common characteristic of both the smart vehicle planning and the intelligent data processing is that their performances can be greatly accelerated using GPU, hence parallelism.
-For example, the more a self-driving car plans its potential paths to follow {\it a priori}, the better the quality of self-driving is.
-Since the self-driving car should follow only the best path, calculation of each potential path can happen parallel.
-A CMU team has already started using GPUs for autonomous driving computations.
-They have increased the performance of motion planning for autonomous driving up to 40 times using an NVIDIA GPU GTX 260 having 192 logical cores~\cite{McNaughton11}.
-
-The perception system which intelligently processes the data from various sensors can also obtain benefits using GPU.
-The perception system of a self-driving car should be able to detect, classify and track the obstacles around itself.
-Various types of sensors will generate huge amount of information to understand the surroundings.
-In order to process the data, a parallel processing technique is of vital importance.
-For example, a new self-driving car at CMU manages 1536 objects from LIDAR sensors before they are fused with other types of sensor data.
-There has been on-going research using GPU to build a perception system~\cite{Ferreira11}, and their GPU implementation yielded 30,000-times-faster performance compared to the case of using only CPU.
-
-As described above, there has been research on applying GPU to different applications on self-driving cars, and it is clear that GPU can provide great benefits on realizing self-driving car technologies.
-However, not much research has been done when those technologies are deployed together on self-driving cars, where the loads of each application dynamically vary depending on environments.
-The period and the computation time of the planning algorithms for autonomous driving highly depend on the vehicle speed, so the planning algorithms can be heavily loaded when the car is on highway.
-The load of the perception algorithms mainly depend on the number of obstacles around the car hence the perception system requires more computing resources when the car is driving in the urban area.
-Therefore, an intelligent method of sharing many cores on GPU would be essential when we use GPU on self-driving cars.
-For example, if a self-driving car has a 96-core GPU, the planning algorithm of the car can use 72 cores on the highway and use 12 cores in the urban environment.
-A self-driving car~\cite{Urmson08} require tens of tasks, and the dynamic core management should be fulfilled across all tasks if those tasks utilize GPU.
-
-In this paper, we propose conceptual frameworks for GPU applications to adjust their task execution times depending on their given workloads.
-
-The rest of paper is organized as follows.
-Section~\ref{sec:system_model} describes how our proposed system is modeled.
-Section~\ref{sec:adaptivity_support} presents the methods for adaptively managing the given GPU resources, and we conclude our paper in Section~\ref{sec:summary}.
 
@@
 
-We assume such embedded systems that equip GPUs as compute devices.
-The application task starts execution on the CPU, and offloads its
-data-parallel compute-intensive workload onto the GPU when needed.
-Once offloaded onto the GPU, the task becomes non-preemptive due to many
-reasons.
-In fact, it is technically possible to preempt the running task on the
-GPU by loading and restoring its context, but it requires additional
-firmware, runtime, and OS support, and the preemption cost would be
-non-trivial due to a very large set of GPU registers and states.
-We hence restrict our attention to a non-preemptive execution model for
-GPU computing.
-GPUs may also pose some constraints in multi-tasking.
-Even the NVIDIA Fermi architecture~\cite{Fermi}, one of the most popular
-GPU product lines, allows only one context to use GPU resources at
-once, though this context may spawn multiple GPU kernels (jobs)
+sporadic/periodic tasks etc.
+NVIDIA Fermi GPUs.
+NVIDIA Kepler GPUs may allow multiple contexts to execute on the GPU
 simultaneously.
-In other words, if task-level parallelism is required, the entire system
-must run in the same context.
-We however believe that this constraint will not limit the concept of
-this paper.
-We can use the same context to exploit concurrent parallel job
-executions, and it should be enough to demonstrate a proof of concept. 
-We also expect that future product lines will remove this concern.
-
-We consider real-time applications where each task runs in a periodic or
-sporadic manner under deadline constraints.
-Such a task set and application includes motion planning and 
-vision-based perception in state-of-the-art autonomous driving
-vehicles, where the periods often correspond to frame-rates, and the
-deadlines appear at the end of period.
-It should however be noted that deadline constraints may not be relevant
-to periods depending on applications, but the computation must meet the
-deadline in order to achieve the desired quality of service in either case.
-We also presume that the computing demand of each task is highly
-variable.
-For example, the performance of planning and perception is usually
-governed by the number of objects, the size of data, and the desired
-quality of output.
-These workloads are also well parallelizable using the GPU.
-Hence, the autonomous driving tasks and applications are our primary
-target, but the contribution of this paper is not limited to those but
-is also generally applicable to highly vaiable workloads running on the
-GPU.
 
@@
 adjust it at runtime.
-It is important to note that programmer is typically responsible for
+It is important to note that programmers are typically responsible for
 allocating the number of cores (or threads mapped to cores) in GPU
@@
 In order to adjust the number of cores at runtime, it is essential to
-provide the programmer with an interface to obtain the information on
-the number of cores available or allocated for the program at runtime.
-The programmer is then responsible for making the program adaptive to
-the number of cores.
-
+provide the programmers with an interface to obtain the information on
+the number of cores available or allocated for the program.
 In the following, we present two frameworks that could achieve our idea.
 Although they are still at an idea stage, we plan to implement a real
-system as a proof-of-concept, leveraging open-source
+system and demonstrate its capability, leveraging open-source
 software~\cite{Kato_OSPERT11}.
@@
 \subsection{Explicit Adjustment}
-
-We propose an explicit adjustment framework where the programmer is
-responsible for adjusting the number of cores to relax or tighten
-the computing demand.
-There will be no adjustment unless the programmer explicitly takes an
-action.
-A typical usage of this framework with periodic real-time tasks is as
-follows:
 
@@
 
-A downside of this framework is that some task may misbehave
-and interfer with other contending application tasks, if the programmer
-fails to call the API functions correctly.
-We can cap the maximum number of cores available for an individual task
-to prevent it from abusing GPU resources, but adaptivity of computing
-depends on programmers, and the system cannot really control it.
-
 \subsection{Implicit Adjustment}
-
-Our second approach to adaptive resource management is an implicit
-adjustment framework.
-In this framework, the number of cores to be allocated for the program
-is set by the runtime system.
-Hence, adaptivity of computing does not really depend on programmers,
-but if the program is not aware of this framework, it may fail to run,
-since the number of core allocated for the program may be different from
-what the program assumes.
 
@@
 before the task starts.
-If this set point is not specified, the runtime system tries to derive
-it internally as time goes by.
 When the task uses the GPU, the runtime system consistently updates the
@@
 
-This implicit adjustment framework is more encouraged than the explicit
-adjustment framework, as it can enforce adaptive GPU resource
-management.
-However, it requires consensus in the programming model that the number
-of cores allocated for the program could be changed every time it is
-offloaded onto the GPU, and the programmer must be aware of it to make
-the program work.
-We claim that this is a natural trade-off between generality of
-programming and adaptivity of service.
-
 \section{Summary}
@@
 
-In this paper, we have discussed adaptivity requirements in embedded
-real-time systems with GPUs, and presented two frameworks for adaptive
-GPU resource management.
-We conjecture that generality of programming may need to be compromised
-to achieve adaptivity of resource allocation on the GPU.
-Nonetheless, adaptive resource management is a key solution in
-optimizing performance under resource-constrained environments.
-We believe that our frameworks for GPU programming are useful
-contributions in this line of work, given that trends are apparently
-shifting to highly parallel heterogeneous computing.
-
+%\setlength{\bibsep}{2.5pt}
+%\renewcommand{\baselinestretch}{0.94}
 \bibliographystyle{plain}

### 2012-03-09 draft/draft.tex
Abstract is added, and the introduction is modified accordingly.

--- 2012-03-09 draft/draft.tex
@@
  In this paper, we present conceptual frameworks for GPU applications to
- adjust their task execution times upon given workloads.
+ adjust their task execution times upon given workloads. We propose two 
+ frameworks enabling smart GPU resource management when many applications 
+ share GPU resources while the workloads of those applications vary. 
+ For {\it explicit adjustment}, we provide API functions so that application 
+ developers can adjust the number of GPU cores depending on their needs. 
+ An {\it implicit adjustment} will provide a run-time framwork which 
+ dynamically allocates the number of cores to tasks based on the given 
+ workloads. The runtime support of the proposed system can be realized using 
+ functions which measure the execution times of the tasks on GPU and change 
+ the number of GPU cores. We motivate the necessity of this framework 
+ in the context of self-driving technologies, and we believe that our 
+ frameworks for GPU programming are useful contributions given that 
+ trends are apparently shifting to highly parallel heterogeneous computing.
 \end{abstract}
@@
 
-In this paper, we propose conceptual frameworks for GPU applications to adjust their task execution times depending on their given workloads.
+In this paper, we present conceptual frameworks for
+GPU applications to adjust their task execution times upon
+given workloads. We propose two frameworks enabling
+smart GPU resource management when many applications
+share GPU resources while the workloads of those appli-
+cations vary. For explicit adjustment, we provide API func-
+tions so that application developers can adjust the number
+of GPU cores depending on their needs. An implicit adjust-
+ment will provide a run-time framwork which dynamically
+allocates the number of cores to tasks based on the given
+workloads. The runtime support of the proposed system
+can be realized using functions which measure the execu-
+tion times of the tasks on GPU and change the number of
+cores.
 
@@
 is set by the runtime system.
-Hence, adaptivity of computing does not really depend on programmers,
-but if the program is not aware of this framework, it may fail to run,
-since the number of core allocated for the program may be different from
-what the program assumes.
+Hence, adaptivity of computing does not really depend on programmers.
+If the program is not aware of this framework, however, it may fail to
+run, since the number of core allocated for the program may be different
+from what the program assumes.
 
@@
 
+\subsection{Runtime System Support}
+\label{sec:runtime}
+
+The runtime system provides the API for programmers.
+In order to support adaptive GPU resource management, we must provide
+some additional API functions.
+
+\begin{itemize}
+ \item Our adaptive GPU resource management frameworks require a
+       function to measure the execution time of each job running on the
+       GPU.
+       This function is easy-to-implement.
+       Since we assume that job execution on the GPU is non-preemptive,
+       the amount of time in run-to-completion can be accounted as a job
+       execution time.
+       This accounting method is known to work in previous
+       work~\cite{Kato_ATC11, Rossbach_SOSP11}, and hence we can adopt
+       it.
+       For the explicit adjustment framework, this function must be
+       exposed to the programmer, while it can also be used internally
+       by the implicit adjustment framework.
+ \item We also need several functions to change the number of cores to
+       be allocated for the program.
+       Some existing programming languages for GPGPU, \textit{e.g.},
+       CUDA~\cite{CUDA}, provide the API to allow the programmer to
+       specify the shape of grid structure and the number of threads
+       mapped to compute cores.
+       We can use this API as it is, or provide a corresponding API if
+       the underlying programming language does not support it.
+\end{itemize}
+
+In addition to these API functions, the runtime system must be able to
+detect when the program is offloaded onto the GPU and when it is
+completed on the GPU.
+Since the programmer calls a specific API function to launch the GPU
+program in most GPU programming models, it is very easy to record the
+start time of GPU execution.
+Detection of the completion time of GPU execution, on the other hand, is
+not straightforward.
+We would need to use an interrupt to notify the runtime system of
+completion of GPU execution.
+Polling on a particular register is an alternative approach, but it
+would not be suitable for latency-sensitive real-time systems, as
+previous work demonstrated~\cite{Kato_ATC11}.
+
+Finally, runtime system support must be integrated with the API so that
+the programmer can make use of our frameworks under a single unified
+programming model.
+We plan to extend our CUDA runtime library developed in previous
+work~\cite{Kato_OSPERT11} to support our adaptive frameworks.
+This is our prototype implementation, and our frameworks can also be
+integrated with other programming models beyond CUDA.
+
 \section{Summary}

### 2012-03-09 draft/draft.tex
The figure is added, and the tex file is also modified.

@@
 \and
-Shinpei Kato and Scott Brandt\\
+Shinpei Kato\\
 \\
@@
 \label{sec:introduction}
-Graphics processing units (GPUs) are widely used in a range of applications from supercomputers to mobile computers.
-For example, three of the top five supercomputers on the TOP500 list~\cite{TOP500} utilize GPUs to accelerate computations, and recent tablets such as ASUS Eee Pad Transformer Prime leverage 12 GPU cores of NVIDIA’s well-known SoC, Tegra 3 to boost the performance while saving power~\cite{Tegra3}.
-This trend will only continue and in fact is expected to accelerate across multiple application domains.
 
-Modern automobiles have several tens of processing units, and advances in safe-driving features such as adaptive cruise control, stop-and-go cruise control, lane keeping and assisted lane change require even higher performance computing.
-As vehicles become semi-autonomous, drive autonomously on demand and eventually become fully autonomous, a multitude of computer vision, sensor fusion, signal processing and graphics problems must be solved in real time.
-Since solutions to those problems can leverage benefits of parallel processing techniques, GPUs can be used to accelerate computations.
-As an initial step, NVIDIA GPUs will be used for infotainment systems in future BMW vehicles~\cite{NVIDIA_BMW}.
+Graphics processing units (GPUs) are getting more and more commonplace
+in many application domains widely ranging from high-performance
+computing to embedded mobile computing.
+For example, three of the top five supercomputers on the TOP500
+list~\cite{TOP500}, announced as of March 2012, use GPUs to accelerate
+computations, while recent tablets, such as ASUS Eee Pad Transformer
+Prime, also leverage embedded GPUs, \textit{a.k.a.}, Tegra
+3~\cite{Tegra3}, to enhance performance under power constraints.
+This trend will only continue and in fact is expected to appear across
+more and more application domains.
 
-Recent trends in self-driving cars~\cite{Urmson08, Markoff10, Kelly12} are motivated by various benefits of automated driving, but most importantly enabling automatic safety features.
-Those features are realized by the smart vehicle planning and the intelligent processing of data from various types of sensors on self-driving cars such as LIDAR (LIght Detection And Ranging), radar, camera, ultrasonic sensor, and so forth.
-One common characteristic of both the smart vehicle planning and the intelligent data processing is that their performances can be greatly accelerated using GPU, hence parallelism.
-For example, the more a self-driving car plans its potential paths to follow {\it a priori}, the better the quality of self-driving is.
-Since the self-driving car should follow only the best path, calculation of each potential path can happen parallel.
-A CMU team has already started using GPUs for autonomous driving computations.
-They have increased the performance of motion planning for autonomous driving up to 40 times using an NVIDIA GPU GTX 260 having 192 logical cores~\cite{McNaughton11}.
+One notable application domain of GPUs is automotive engineering.
+Modern automobiles employ several tens of processing units.
+Further advances in safe-driving features, such as adaptive
+cruise control, stop-and-go cruise control, lane keeping, and assisted
+lane change, would require even higher computing capabilities.
+As vehicles are becoming autonomous, a multitude of computer vision,
+sensor fusion, signal processing, and graphics systems must operate and
+communicate in real-time.
+Given their highly data-parallel and compute-intensive workloads, 
+we believe that parallel computing is a useful solution.
+As technology reads today, the GPU is the most well-suited platform.
+In fact, NVIDIA GPUs will be used for infotainment systems platforms in
+future product lines of BMW vehicles~\cite{NVIDIA_BMW}.
 
-The perception system which intelligently processes the data from various sensors can also obtain benefits using GPU.
-The perception system of a self-driving car should be able to detect, classify and track the obstacles around itself.
-Various types of sensors will generate huge amount of information to understand the surroundings.
-In order to process the data, a parallel processing technique is of vital importance.
-For example, a new self-driving car at CMU manages 1536 objects from LIDAR sensors before they are fused with other types of sensor data.
-There has been on-going research using GPU to build a perception system~\cite{Ferreira11}, and their GPU implementation yielded 30,000-times-faster performance compared to the case of using only CPU.
+Recent trends in autonomous-driving vehicles~\cite{Kelly12, Markoff10,
+Urmson08} are motivated by various benefits of automated functions,
+but, most importantly, enabling automatic safety features. 
+These features are often brought by smart planning and intelligent
+processing of data obtained from many sensors equiped in the vehicle,
+including LIDAR (LIght Detection And Ranging), radar, camera, and
+ultrasonic sensors. 
+One common characteristic in these types of processing is that GPUs can
+accelerate their processing speeds significantly.
+For example, autonomous driving should follow only the best path among
+many potential paths whose calculations can happen in parallel.
+Calculating as many potential paths to follow as possible
+{\it a priori}, we will get better quality of driving.
+As a matter of fact, CMU's autonomous vehicle team showed that their
+motion planning algorithm was sped up by 40 times~\cite{McNaughton11},
+using an NVIDIA GTX 260 GPU that integrates 192 compute cores on a
+chip.
+
+In addition to motion planning, a perception algorithm can benefit
+from the GPU as well in sensor data processing.
+The perception system of a self-driving car should
+be able to detect, classify and track the obstacles around
+itself. Various types of sensors will generate huge amount of
+information to understand the surroundings. In order to process the
+data, a parallel processing technique is of vital importance. For
+example, a new self-driving car at CMU manages 1536 objects from LIDAR
+sensors before they are fused with other types of sensor data. There has
+been on-going research using GPU to build a perception
+system~\cite{Ferreira11}, and their GPU implementation yielded
+30,000-times-faster performance compared to the case of using only CPU.
 
@@
 
+\begin{figure}[t]
+\includegraphics[width=3.0in]{architecture.eps}
+\caption{The proposed conceptual architecture for GPU resource management.} 
+\label{fig_arch}
+\end{figure}
+
 In the following, we present two frameworks that could achieve our idea.
@@
 system as a proof-of-concept, leveraging open-source
-software~\cite{Kato_OSPERT11}.
+software~\cite{Kato_OSPERT11}. The proposed architecture is also 
+illustrated in Figure~\ref{fig_arch}.
 

### 2012-03-09 draft/draft.tex
added Section 2 (System Model)

--- 2012-03-09 draft/draft.tex
@@
 
-sporadic/periodic tasks etc.
-NVIDIA Fermi GPUs.
-NVIDIA Kepler GPUs may allow multiple contexts to execute on the GPU
+We assume such embedded systems that equip GPUs as compute devices.
+The application task starts execution on the CPU, and offloads its
+data-parallel compute-intensive workload onto the GPU when needed.
+Once offloaded onto the GPU, the task becomes non-preemptive due to many
+reasons.
+In fact, it is technically possible to preempt the running task on the
+GPU by loading and restoring its context, but it requires additional
+firmware, runtime, and OS support, and the preemption cost would be
+non-trivial due to a very large set of GPU registers and states.
+We hence restrict our attention to a non-preemptive execution model for
+GPU computing.
+GPUs may also pose some constraints in multi-tasking.
+Even the NVIDIA Fermi architecture~\cite{Fermi}, one of the most popular
+GPU product lines, allows only one context to use GPU resources at
+once, though this context may spawn multiple GPU kernels (jobs)
 simultaneously.
+In other words, if task-level parallelism is required, the entire system
+must run in the same context.
+We however believe that this constraint will not limit the concept of
+this paper.
+We can use the same context to exploit concurrent parallel job
+executions, and it should be enough to demonstrate a proof of concept. 
+We also expect that future product lines will remove this concern.
+
+We consider real-time applications where each task runs in a periodic or
+sporadic manner under deadline constraints.
+Such a task set and application includes motion planning and 
+vision-based perception in state-of-the-art autonomous driving
+vehicles.
+We also presume the computing demand of each task to be highly
+variable.
+For example, planning and perception workloads are governed by the
+number of objects, the size of data, and the desired quality of output.
+These workloads are also well parallelizable using the GPU.
+Hence, the autonomous driving tasks and applications are our primary
+target, but the contribution of this paper is not limited to those but
+is also generally applicable to highly vaiable workloads running on the
+GPU.
 

### 2012-03-09 draft/draft.tex
minor update to Section 2

@@
 vision-based perception in state-of-the-art autonomous driving
-vehicles.
-We also presume the computing demand of each task to be highly
+vehicles, where the periods often correspond to frame-rates, and the
+deadlines appear at the end of period.
+It should however be noted that deadline constraints may not be relevant
+to periods depending on applications, but the computation must meet the
+deadline in order to achieve the desired quality of service in either case.
+We also presume that the computing demand of each task is highly
 variable.
-For example, planning and perception workloads are governed by the
-number of objects, the size of data, and the desired quality of output.
+For example, the performance of planning and perception is usually
+governed by the number of objects, the size of data, and the desired
+quality of output.
 These workloads are also well parallelizable using the GPU.

### 2012-03-09 draft/draft.tex
added texts to Section 3

@@
 adjust it at runtime.
-It is important to note that programmers are typically responsible for
+It is important to note that programmer is typically responsible for
 allocating the number of cores (or threads mapped to cores) in GPU
@@
 In order to adjust the number of cores at runtime, it is essential to
-provide the programmers with an interface to obtain the information on
-the number of cores available or allocated for the program.
+provide the programmer with an interface to obtain the information on
+the number of cores available or allocated for the program at runtime.
+The programmer is then responsible for making the program adaptive to
+the number of cores.
+
 In the following, we present two frameworks that could achieve our idea.
 Although they are still at an idea stage, we plan to implement a real
-system and demonstrate its capability, leveraging open-source
+system as a proof-of-concept, leveraging open-source
 software~\cite{Kato_OSPERT11}.
@@
 \subsection{Explicit Adjustment}
+
+We propose an explicit adjustment framework where the programmer is
+responsible for adjusting the number of cores to relax or tighten
+the computing demand.
+There will be no adjustment unless the programmer explicitly takes an
+action.
+A typical usage of this framework with periodic real-time tasks is as
+follows:
 
@@
 
+A downside of this framework is that some task may misbehave
+and interfer with other contending application tasks, if the programmer
+fails to call the API functions correctly.
+We can cap the maximum number of cores available for an individual task
+to prevent it from abusing GPU resources, but adaptivity of computing
+depends on programmers, and the system cannot really control it.
+
 \subsection{Implicit Adjustment}
+
+Our second approach to adaptive resource management is an implicit
+adjustment framework.
+In this framework, the number of cores to be allocated for the program
+is set by the runtime system.
+Hence, adaptivity of computing does not really depend on programmers,
+but if the program is not aware of this framework, it may fail to run,
+since the number of core allocated for the program may be different from
+what the program assumes.
 
@@
 before the task starts.
+If this set point is not specified, the runtime system tries to derive
+it internally as time goes by.
 When the task uses the GPU, the runtime system consistently updates the
@@
 offloading the computation onto the GPU.
+
+This implicit adjustment framework is more encouraged than the explicit
+adjustment framework, as it can enforce adaptive GPU resource
+management.
+However, it requires consensus in the programming model that the number
+of cores allocated for the program could be changed every time it is
+offloaded onto the GPU, and the programmer must be aware of it to make
+the program work.
+We claim that this is a natural trade-off between generality of
+programming and adaptivity of service.
 

### 2012-03-09 draft/draft.tex
added texts to Section 4

@@
 
-%\setlength{\bibsep}{2.5pt}
-%\renewcommand{\baselinestretch}{0.94}
+In this paper, we have discussed adaptivity requirements in embedded
+real-time systems with GPUs, and presented two frameworks for adaptive
+GPU resource management.
+We conjecture that generality of programming may need to be compromised
+to achieve adaptivity of resource allocation on the GPU.
+Nonetheless, adaptive resource management is a key solution in
+optimizing performance under resource-constrained environments.
+We believe that our frameworks for GPU programming are useful
+contributions in this line of work, given that trends are apparently
+shifting to highly parallel heterogeneous computing.
+
 \bibliographystyle{plain}

### 2012-03-10 draft/draft.tex

@@
 \label{sec:introduction}
+Graphics processing units (GPUs) are widely used in a range of applications from supercomputers to mobile computers.
+For example, three of the top five supercomputers on the TOP500 list~\cite{TOP500} utilize GPUs to accelerate computations, and recent tablets such as ASUS Eee Pad Transformer Prime leverage 12 GPU cores of NVIDIA’s well-known SoC, Tegra 3 to boost the performance while saving power~\cite{Tegra3}.
+This trend will only continue and in fact is expected to accelerate across multiple application domains.
+
+Modern automobiles have several tens of processing units, and advances in safe-driving features such as adaptive cruise control, stop-and-go cruise control, lane keeping and assisted lane change require even higher performance computing.
+As vehicles become semi-autonomous, drive autonomously on demand and eventually become fully autonomous, a multitude of computer vision, sensor fusion, signal processing and graphics problems must be solved in real time.
+Since solutions to those problems can leverage benefits of parallel processing techniques, GPUs can be used to accelerate computations.
+As an initial step, NVIDIA GPUs will be used for infotainment systems in future BMW vehicles~\cite{NVIDIA_BMW}.
+
+Recent trends in self-driving cars~\cite{Urmson08, Markoff10, Kelly12} are motivated by various benefits of automated driving, but most importantly enabling automatic safety features.
+Those features are realized by the smart vehicle planning and the intelligent processing of data from various types of sensors on self-driving cars such as LIDAR (LIght Detection And Ranging), radar, camera, ultrasonic sensor, and so forth.
+One common characteristic of both the smart vehicle planning and the intelligent data processing is that their performances can be greatly accelerated using GPU, hence parallelism.
+For example, the more a self-driving car plans its potential paths to follow {\it a priori}, the better the quality of self-driving is.
+Since the self-driving car should follow only the best path, calculation of each potential path can happen parallel.
+A CMU team has already started using GPUs for autonomous driving computations.
+They have increased the performance of motion planning for autonomous driving up to 40 times using an NVIDIA GPU GTX 260 having 192 logical cores~\cite{McNaughton11}.
+
+The perception system which intelligently processes the data from various sensors can also obtain benefits using GPU.
+The perception system of a self-driving car should be able to detect, classify and track the obstacles around itself.
+Various types of sensors will generate huge amount of information to understand the surroundings.
+In order to process the data, a parallel processing technique is of vital importance.
+For example, a new self-driving car at CMU manages 1536 objects from LIDAR sensors before they are fused with other types of sensor data.
+There has been on-going research using GPU to build a perception system~\cite{Ferreira11}, and their GPU implementation yielded 30,000-times-faster performance compared to the case of using only CPU.
+
+As described above, there has been research on applying GPU to different applications on self-driving cars, and it is clear that GPU can provide great benefits on realizing self-driving car technologies.
+However, not much research has been done when those technologies are deployed together on self-driving cars, where the loads of each application dynamically vary depending on environments.
+The period and the computation time of the planning algorithms for autonomous driving highly depend on the vehicle speed, so the planning algorithms can be heavily loaded when the car is on highway.
+The load of the perception algorithms mainly depend on the number of obstacles around the car hence the perception system requires more computing resources when the car is driving in the urban area.
+Therefore, an intelligent method of sharing many cores on GPU would be essential when we use GPU on self-driving cars.
+For example, if a self-driving car has a 96-core GPU, the planning algorithm of the car can use 72 cores on the highway and use 12 cores in the urban environment.
+A self-driving car~\cite{Urmson08} require tens of tasks, and the dynamic core management should be fulfilled across all tasks if those tasks utilize GPU.
+
+In this paper, we propose conceptual frameworks for GPU applications to adjust their task execution times depending on their given workloads.
+
+The rest of paper is organized as follows.
+Section~\ref{sec:system_model} describes how our proposed system is modeled.
+Section~\ref{sec:adaptivity_support} presents the methods for adaptively managing the given GPU resources, and we conclude our paper in Section~\ref{sec:summary}.
 

### 2012-03-10 draft/draft.tex
some edits to Introduction

--- 2012-03-10 draft/draft.tex
@@
 \label{sec:introduction}
+
+Graphics processing units (GPUs) are getting more and more commonplace
+in many application domains widely ranging from high-performance
+computing to embedded mobile computing.
+For example, three of the top five supercomputers on the TOP500
+list~\cite{TOP500}, announced as of March 2012, use GPUs to accelerate
+computations, while recent tablets, such as ASUS Eee Pad Transformer
+Prime, also leverage embedded GPUs, \textit{a.k.a.}, Tegra
+3~\cite{Tegra3}, to enhance performance under power constraints.
+This trend will only continue and in fact is expected to appear across
+more and more application domains.
+
+One notable application domain of GPUs is automotive engineering.
+Modern automobiles employ several tens of processing units.
+Further advances in safe-driving features, such as adaptive
+cruise control, stop-and-go cruise control, lane keeping, and assisted
+lane change, would require even higher computing capabilities.
+As vehicles are becoming autonomous, a multitude of computer vision,
+sensor fusion, signal processing, and graphics systems must operate and
+communicate in real-time.
+Given their highly data-parallel and compute-intensive workloads, 
+we believe that parallel computing is a useful solution.
+As technology reads today, the GPU is the most well-suited platform.
+In fact, NVIDIA GPUs will be used for infotainment systems platforms in
+future product lines of BMW vehicles~\cite{NVIDIA_BMW}.
+
+Recent trends in autonomous-driving vehicles~\cite{Kelly12, Markoff10,
+Urmson08} are motivated by various benefits of automated functions,
+but, most importantly, enabling automatic safety features. 
+These features are often brought by smart planning and intelligent
+processing of data obtained from many sensors equiped in the vehicle,
+including LIDAR (LIght Detection And Ranging), radar, camera, and
+ultrasonic sensors. 
+One common characteristic in these types of processing is that GPUs can
+accelerate their processing speeds significantly.
+For example, autonomous driving should follow only the best path among
+many potential paths whose calculations can happen in parallel.
+Calculating as many potential paths to follow as possible
+{\it a priori}, we will get better quality of driving.
+As a matter of fact, CMU's autonomous vehicle team showed that their
+motion planning algorithm was sped up by 40 times~\cite{McNaughton11},
+using an NVIDIA GTX 260 GPU that integrates 192 compute cores on a
+chip.
+
+In addition to motion planning, a perception algorithm can benefit
+from the GPU as well in sensor data processing.
+The perception system of a self-driving car should
+be able to detect, classify and track the obstacles around
+itself. Various types of sensors will generate huge amount of
+information to understand the surroundings. In order to process the
+data, a parallel processing technique is of vital importance. For
+example, a new self-driving car at CMU manages 1536 objects from LIDAR
+sensors before they are fused with other types of sensor data. There has
+been on-going research using GPU to build a perception
+system~\cite{Ferreira11}, and their GPU implementation yielded
+30,000-times-faster performance compared to the case of using only CPU.
+
+As described above, there has been research on applying GPU to different applications on self-driving cars, and it is clear that GPU can provide great benefits on realizing self-driving car technologies.
+However, not much research has been done when those technologies are deployed together on self-driving cars, where the loads of each application dynamically vary depending on environments.
+The period and the computation time of the planning algorithms for autonomous driving highly depend on the vehicle speed, so the planning algorithms can be heavily loaded when the car is on highway.
+The load of the perception algorithms mainly depend on the number of obstacles around the car hence the perception system requires more computing resources when the car is driving in the urban area.
+Therefore, an intelligent method of sharing many cores on GPU would be essential when we use GPU on self-driving cars.
+For example, if a self-driving car has a 96-core GPU, the planning algorithm of the car can use 72 cores on the highway and use 12 cores in the urban environment.
+A self-driving car~\cite{Urmson08} require tens of tasks, and the dynamic core management should be fulfilled across all tasks if those tasks utilize GPU.
+
+In this paper, we propose conceptual frameworks for GPU applications to adjust their task execution times depending on their given workloads.
+
+The rest of paper is organized as follows.
+Section~\ref{sec:system_model} describes how our proposed system is modeled.
+Section~\ref{sec:adaptivity_support} presents the methods for adaptively managing the given GPU resources, and we conclude our paper in Section~\ref{sec:summary}.
 

### 2012-03-14 draft/draft.tex
I reflected Raj's comments.

--- 2012-03-10 draft/draft.tex
@@
 \title{
-Toward Adaptive GPU Resource Management for Embedded Real-Time Systems
+Towards Adaptive GPU Resource Management for Embedded Real-Time Systems
 }
@@
 \begin{abstract}
- In this paper, we present conceptual frameworks for GPU applications to
- adjust their task execution times upon given workloads.
+ In this paper, we present two conceptual frameworks for GPU applications to
+ adjust their task execution times based on total workload. These  
+ frameworks enable smart GPU resource management when many applications 
+ share GPU resources while the workloads of those applications vary. 
+ Application 
+ developers can explicitly adjust the number of GPU cores depending on their needs. 
+ An implicit adjustment will be supported by a run-time framwork, which 
+ dynamically allocates the number of cores to tasks based on the total 
+ workload. The runtime support of the proposed system can be realized using 
+ functions which measure the execution times of the tasks on GPU and change 
+ the number of GPU cores. We motivate the necessity of this framework 
+ in the context of self-driving technologies, and we believe that our 
+ frameworks for GPU programming are useful contributions given 
+ the increasing emphasis on 
+ %trends are apparently shifting to highly 
+ parallel heterogeneous computing.
 \end{abstract}
@@
 
-Graphics processing units (GPUs) are getting more and more commonplace
+Graphics processing units (GPUs) are becoming more and more commonplace
 in many application domains widely ranging from high-performance
@@
 computations, while recent tablets, such as ASUS Eee Pad Transformer
-Prime, also leverage embedded GPUs, \textit{a.k.a.}, Tegra
+Prime, also leverage embedded GPUs, like Tegra
 3~\cite{Tegra3}, to enhance performance under power constraints.
-This trend will only continue and in fact is expected to appear across
-more and more application domains.
+This trend is expected to continue.
 
@@
 cruise control, stop-and-go cruise control, lane keeping, and assisted
-lane change, would require even higher computing capabilities.
-As vehicles are becoming autonomous, a multitude of computer vision,
-sensor fusion, signal processing, and graphics systems must operate and
+lane change, would require even larger computing capabilities.
+%As vehicles are becoming autonomous, 
+For vehicles to become fully or semi-autonomous, 
+a multitude of computer vision,
+sensor fusion, signal processing, and graphics sub-systems must operate and
 communicate in real-time.
 Given their highly data-parallel and compute-intensive workloads, 
-we believe that parallel computing is a useful solution.
-As technology reads today, the GPU is the most well-suited platform.
+parallel computing is a useful solution.
+As technology stands today, the GPU is the most well-suited platform.
 In fact, NVIDIA GPUs will be used for infotainment systems platforms in
@@
 
-Recent trends in autonomous-driving vehicles~\cite{Kelly12, Markoff10,
-Urmson08} are motivated by various benefits of automated functions,
-but, most importantly, enabling automatic safety features. 
-These features are often brought by smart planning and intelligent
+%Recent trends in autonomous-driving vehicles~\cite{Kelly12, Markoff10,
+%Urmson08} are motivated by various benefits of automated functions,
+%but, most importantly, enabling 
+Automatic safety features require smart planning and intelligent
 processing of data obtained from many sensors equiped in the vehicle,
@@
 accelerate their processing speeds significantly.
-For example, autonomous driving should follow only the best path among
-many potential paths whose calculations can happen in parallel.
+For example, autonomous driving should ideally follow the best path among
+many potential paths, whose calculations can happen in parallel.
 Calculating as many potential paths to follow as possible
-{\it a priori}, we will get better quality of driving.
+will yield better quality of driving.
 As a matter of fact, CMU's autonomous vehicle team showed that their
@@
 
-In addition to motion planning, a perception algorithm can benefit
-from the GPU as well in sensor data processing.
+In addition to motion planning, a perception algorithm as well as 
+sensor data processing can benefit from the GPU.
 The perception system of a self-driving car should
 be able to detect, classify and track the obstacles around
-itself. Various types of sensors will generate huge amount of
-information to understand the surroundings. In order to process the
-data, a parallel processing technique is of vital importance. For
-example, a new self-driving car at CMU manages 1536 objects from LIDAR
+itself. Various types of sensors will generate voluminous amount of
+information that must be processed in order to understand the vehicle's surroundings. 
+%In order to process the data, a parallel processing technique is of vital importance. 
+For example, a self-driving car at CMU manages 1536 objects from LIDAR
 sensors before they are fused with other types of sensor data. There has
@@
 system~\cite{Ferreira11}, and their GPU implementation yielded
-30,000-times-faster performance compared to the case of using only CPU.
+30,000-times-faster performance compared to the case of using only one CPU.
 
-As described above, there has been research on applying GPU to different applications on self-driving cars, and it is clear that GPU can provide great benefits on realizing self-driving car technologies.
-However, not much research has been done when those technologies are deployed together on self-driving cars, where the loads of each application dynamically vary depending on environments.
-The period and the computation time of the planning algorithms for autonomous driving highly depend on the vehicle speed, so the planning algorithms can be heavily loaded when the car is on highway.
-The load of the perception algorithms mainly depend on the number of obstacles around the car hence the perception system requires more computing resources when the car is driving in the urban area.
-Therefore, an intelligent method of sharing many cores on GPU would be essential when we use GPU on self-driving cars.
+As described above, there has been research on applying GPU to different applications on self-driving cars, and it is clear that GPU can provide great benefits on realizing safer or self-driving car technologies.
+However, not much research has been done when those technologies are deployed together on self-driving cars, where the loads of each application dynamically vary depending on the environment.
+The period and the computation time of the planning algorithms for autonomous driving highly depend on the vehicle speed, so the planning algorithms can be heavily loaded when the car is (say) on a highway.
+The load of the perception algorithms mainly depends on the number of obstacles around the car.
+Hence, the perception system requires more computing resources when the car is driving in an urban area.
+Therefore, an intelligent method of sharing many cores on GPU would be essential when we use GPUs on self-driving cars.
 For example, if a self-driving car has a 96-core GPU, the planning algorithm of the car can use 72 cores on the highway and use 12 cores in the urban environment.
-A self-driving car~\cite{Urmson08} require tens of tasks, and the dynamic core management should be fulfilled across all tasks if those tasks utilize GPU.
+A self-driving car~\cite{Urmson08} requires tens of tasks, and the dynamic core management should be fulfilled across all tasks if those tasks utilize GPU.
 
-In this paper, we propose conceptual frameworks for GPU applications to adjust their task execution times depending on their given workloads.
+In this paper, we present two conceptual frameworks for
+GPU applications to adjust their task execution times 
+given current workload conditions. These frameworks enable
+smart GPU resource management when many applications
+share GPU resources while the workloads of those applications vary. 
+These frameworks support both explicit and implicit adjustment. 
+With support for explicit adjustment, %we provide API functions so that 
+application developers can adjust the number
+of GPU cores depending on their needs. %An implicit adjustment will provide 
+A run-time framwork will dynamically
+allocate the number of cores to tasks based on current 
+workloads. The runtime support of our proposed system
+can be realized using functions which measure the execution 
+times of the tasks on GPU and change the number of cores.
 
@@
 Section~\ref{sec:system_model} describes how our proposed system is modeled.
-Section~\ref{sec:adaptivity_support} presents the methods for adaptively managing the given GPU resources, and we conclude our paper in Section~\ref{sec:summary}.
+Section~\ref{sec:adaptivity_support} presents the methods for adaptively managing GPU resources, and we conclude our paper in Section~\ref{sec:summary}.
 
@@
 
-We assume such embedded systems that equip GPUs as compute devices.
-The application task starts execution on the CPU, and offloads its
+We assume real-time embedded systems that contain CPU and one or more GPUs as compute devices.
+An application task starts execution on the CPU, and offloads its
 data-parallel compute-intensive workload onto the GPU when needed.
@@
 non-trivial due to a very large set of GPU registers and states.
-We hence restrict our attention to a non-preemptive execution model for
+We, hence, restrict our attention to a non-preemptive execution model for
 GPU computing.
@@
 must run in the same context.
-We however believe that this constraint will not limit the concept of
-this paper.
-We can use the same context to exploit concurrent parallel job
-executions, and it should be enough to demonstrate a proof of concept. 
+We, however, believe that this constraint will not limit the 
+concepts we describe in this paper.
+The same GPU context can be used to exploit concurrent parallel job
+executions, to serve at least as a proof of concept. 
 We also expect that future product lines will remove this concern.
@@
 sporadic manner under deadline constraints.
-Such a task set and application includes motion planning and 
+Such a task set may include motion planning and 
 vision-based perception in state-of-the-art autonomous driving
 vehicles, where the periods often correspond to frame-rates, and the
-deadlines appear at the end of period.
-It should however be noted that deadline constraints may not be relevant
-to periods depending on applications, but the computation must meet the
-deadline in order to achieve the desired quality of service in either case.
+deadlines occur at the end of the period.
+%It should however be noted that deadline constraints may not be relevant
+%to periods depending on applications, but the computation must meet the
+%deadline in order to achieve the desired quality of service in either case.
 We also presume that the computing demand of each task is highly
 variable.
-For example, the performance of planning and perception is usually
+For example, the performance of planning and perception tasks is usually
 governed by the number of objects, the size of data, and the desired
 quality of output.
-These workloads are also well parallelizable using the GPU.
-Hence, the autonomous driving tasks and applications are our primary
-target, but the contribution of this paper is not limited to those but
-is also generally applicable to highly vaiable workloads running on the
+These workloads are also very parallelizable using the GPU.
+%Hence, the autonomous driving tasks and applications are our primary
+%target, but 
+The contributions of this paper are not limited to autonomous driving tasks but
+are also generally applicable to highly variable workloads running on the
 GPU.
@@
 
-In this section, we provide an idea of adaptivity support for GPU
+In this section, we describe adaptivity support for GPU
 applications.
-We particularly focus on computing resource allocation problems.
-It is true that some embedded real-time systems exhibit highly variable
+We particularly focus on solving resource allocation problems.
+The goal is to support embedded real-time systems that exhibit highly variable
 workloads.
-Since GPUs integrate a great number of compute cores on a chip, we can
-support even highly variable workloads in a timely manner by adjusting
+Since GPUs integrate a large number of cores on a chip, 
+we aim to enable the execution of 
+highly variable workloads in a timely manner by adjusting
 allocated cores at runtime.
 
-There are several levels of approaches for adaptive GPU resource
+Several approaches have been studied for adaptive GPU resource
 management.
-Previous work~\cite{Kato_RTAS11, Kato_RTSS11, Kato_ATC11} took
-time-driven approaches that control timings and the amounts of time
+Some work~\cite{Kato_RTAS11, Kato_RTSS11, Kato_ATC11} took
+time-driven approach that controls timings and the duration of time
 allowed to access GPU resources, \textit{i.e.}, scheduling and
@@
 by the OS or runtime scheduler.
-However, they can never manage task execution times.
+However, they can not manage task execution times.
 They also limit the number of contexts that can access the GPU
@@
 We consider a different approach than previous work that enables GPU
-applications to adjust their task execution times upon workload.
-We believe that the number of cores used in the program is a major
-factor that could change the execution time, and hence explore how to
-adjust it at runtime.
-It is important to note that programmer is typically responsible for
+applications to adjust their task execution times.
+The number of cores used in the program is a major
+factor that affects the execution time. Hence, we explore how to
+adjust the core allocation at runtime.
+It is important to note that the programmer is typically responsible for
 allocating the number of cores (or threads mapped to cores) in GPU
@@
 
-In the following, we present two frameworks that could achieve our idea.
-Although they are still at an idea stage, we plan to implement a real
+\begin{figure}[t]
+\includegraphics[width=3.0in]{architecture.eps}
+\caption{The proposed conceptual architecture for GPU resource management.} 
+\label{fig_arch}
+\end{figure}
+
+In the following, we present two frameworks that could be used to 
+implement the proposed approach. 
+%Although they are still at an idea stage, 
+We plan to implement a real
 system as a proof-of-concept, leveraging open-source
-software~\cite{Kato_OSPERT11}.
+software~\cite{Kato_OSPERT11}. The proposed architecture is also 
+illustrated in Figure~\ref{fig_arch}.
 
@@
 
-We propose an explicit adjustment framework where the programmer is
+In our explicit adjustment framework, the programmer is
 responsible for adjusting the number of cores to relax or tighten
@@
 A typical usage of this framework with periodic real-time tasks is as
-follows:
+follows.
 
-At the end of each period, the programmer calls a API function provided
+At the end of each period, the programmer calls a function provided
 by our framework that returns the latest task execution time.
 The programmer next calls either of the following two API functions.
-One increases the number of cores to be used in the next period or GPU
-invocation to speed up the program.
-The other decreases it to speed down the program.
-This framework would work fine because the programmer often knows the
+One increases the number of cores to be used by the next GPU execution 
+to speed up the program.
+The other decreases it to slow down the program.
+This framework is usable in practice because the programmer often knows the
 desired task execution time to meet the frame-rate or deadline.
@@
 
-A downside of this framework is that some task may misbehave
-and interfer with other contending application tasks, if the programmer
+A downside of this framework is that a task may misbehave
+and interfere with other contending application tasks, if the programmer
 fails to call the API functions correctly.
 We can cap the maximum number of cores available for an individual task
-to prevent it from abusing GPU resources, but adaptivity of computing
-depends on programmers, and the system cannot really control it.
+to prevent it from abusing GPU resources, but the adaptivity of computing
+depends on the programmer, and outside system control.
 
@@
 is set by the runtime system.
-Hence, adaptivity of computing does not really depend on programmers.
+Hence, the adaptivity of computing does not really depend on the programmer.
 If the program is not aware of this framework, however, it may fail to
@@
 based on the previous execution time records.
-It is a programmer's duty to check the number of available cores before
+It is still the programmer's duty to check the number of available cores before
 offloading the computation onto the GPU.
 
-This implicit adjustment framework is more encouraged than the explicit
+This implicit adjustment framework is more preferable to the explicit
 adjustment framework, as it can enforce adaptive GPU resource
@@
 the program work.
-We claim that this is a natural trade-off between generality of
-programming and adaptivity of service.
+We claim that this is a natural trade-off between the generality of
+programming and needed adaptivity of service.
 
@@
 
-The runtime system provides the API for programmers.
+The runtime system provides the API for real-time GPU programmers.
 In order to support adaptive GPU resource management, we must provide
@@
        GPU.
-       This function is easy-to-implement.
+       This function is easy to implement.
        Since we assume that job execution on the GPU is non-preemptive,
-       the amount of time in run-to-completion can be accounted as a job
+       the amount of time in run-to-completion can be accounted as job
        execution time.
-       This accounting method is known to work in previous
-       work~\cite{Kato_ATC11, Rossbach_SOSP11}, and hence we can adopt
-       it.
+       This accounting method is also known to work from previous
+       studies~\cite{Kato_ATC11, Rossbach_SOSP11}.
        For the explicit adjustment framework, this function must be
-       exposed to the programmer, while it can also be used internally
+       exposed to the programmer, while it is used internally
        by the implicit adjustment framework.
@@
        CUDA~\cite{CUDA}, provide the API to allow the programmer to
-       specify the shape of grid structure and the number of threads
+       specify the shape of the grid structure and the number of threads
        mapped to compute cores.
@@
 start time of GPU execution.
-Detection of the completion time of GPU execution, on the other hand, is
+The detection of the completion time of GPU execution, on the other hand, is
 not straightforward.
 We would need to use an interrupt to notify the runtime system of
-completion of GPU execution.
-Polling on a particular register is an alternative approach, but it
+the completion of GPU execution.
+Polling on a particular register is an alternative, but it
 would not be suitable for latency-sensitive real-time systems, as
@@
 work~\cite{Kato_OSPERT11} to support our adaptive frameworks.
-This is our prototype implementation, and our frameworks can also be
+While this is our planned prototype implementation, and our frameworks can also be
 integrated with other programming models beyond CUDA.
@@
 GPU resource management.
-We conjecture that generality of programming may need to be compromised
+We conjecture that the generality of programming may need to be compromised
 to achieve adaptivity of resource allocation on the GPU.
@@
 We believe that our frameworks for GPU programming are useful
-contributions in this line of work, given that trends are apparently
-shifting to highly parallel heterogeneous computing.
+contributions in this line of work, given the increasing emphasis in 
+highly parallel heterogeneous computing.
 

### 2012-03-15 draft/draft.tex
recovered the citations related to self-driving cars

@@
 sensor fusion, signal processing, and graphics sub-systems must operate and
-communicate in real-time.
+communicate in real-time~\cite{Kelly12, Markoff10, Urmson08}.
 Given their highly data-parallel and compute-intensive workloads, 

### 2012-10-01 draft/draft.tex
updated for SIGBED

@@
-\documentclass[times, 10pt, twocolumn]{article}
-\usepackage{latex8}
-\usepackage[dvips]{graphicx}
-\usepackage[fleqn]{amsmath}
-\usepackage{amsthm}
-\usepackage{txfonts}
+% THIS IS SIGPROC-SP.TEX - VERSION 3.1
+% WORKS WITH V3.2SP OF ACM_PROC_ARTICLE-SP.CLS
+% APRIL 2009
+%
+% It is an example file showing how to use the 'acm_proc_article-sp.cls' V3.2SP
+% LaTeX2e document class file for Conference Proceedings submissions.
+% ----------------------------------------------------------------------------------------------------------------
+% This .tex file (and associated .cls V3.2SP) *DOES NOT* produce:
+%       1) The Permission Statement
+%       2) The Conference (location) Info information
+%       3) The Copyright Line with ACM data
+%       4) Page numbering
+% ---------------------------------------------------------------------------------------------------------------
+% It is an example which *does* use the .bib file (from which the .bbl file
+% is produced).
+% REMEMBER HOWEVER: After having produced the .bbl file,
+% and prior to final submission,
+% you need to 'insert'  your .bbl file into your source .tex file so as to provide
+% ONE 'self-contained' source file.
+%
+% Questions regarding SIGS should be sent to
+% Adrienne Griscti ---> griscti@acm.org
+%
+% Questions/suggestions regarding the guidelines, .tex and .cls files, etc. to
+% Gerald Murray ---> murray@hq.acm.org
+%
+% For tracking purposes - this is V3.1SP - APRIL 2009
+
+\documentclass{acm_proc_article-sp}
+%\usepackage[dvips]{graphicx}
+%\usepackage[fleqn]{amsmath}
+%\usepackage{amsthm}
+%\usepackage{txfonts}
 \usepackage{courier}
@@
 }
+%
+% You need the command \numberofauthors to handle the 'placement
+% and alignment' of the authors beneath the title.
+%
+% For aesthetic reasons, we recommend 'three authors at a time'
+% i.e. three 'name/affiliation blocks' be placed beneath the title.
+%
+% NOTE: You are NOT restricted in how many 'rows' of
+% "name/affiliations" may appear. We just ask that you restrict
+% the number of 'columns' to three.
+%
+% Because of the available 'opening page real-estate'
+% we ask you to refrain from putting more than six authors
+% (two rows with three columns) beneath the article title.
+% More than six makes the first-page appear very cluttered indeed.
+%
+% Use the \alignauthor commands to handle the names
+% and affiliations for an 'aesthetic maximum' of six authors.
+% Add names, affiliations, addresses for
+% the seventh etc. author(s) as the argument for the
+% \additionalauthors command.
+% These 'additional authors' will be output/set for you
+% without further effort on your part as the last section in
+% the body of your article BEFORE References or any Appendices.
 
-\author {
-Junsung Kim and Ragunathan (Raj) Rajkumar\\
-\\
-Department of Electrical and Computer Engineering\\
-Carnegie Mellon University
+\numberofauthors{2} %  in this sample file, there are a *total*
+% of EIGHT authors. SIX appear on the 'first-page' (for formatting
+% reasons) and the remaining two appear in the \additionalauthors section.
+%
+\author{
+% You can go ahead and credit any number of authors here,
+% e.g. one 'row of three' or two rows (consisting of one row of three
+% and a second row of one, two or three).
+%
+% The command \alignauthor (no curly braces needed) should
+% precede each author name, affiliation/snail-mail address and
+% e-mail address. Additionally, tag each line of
+% affiliation/address with \affaddr, and tag the
+% e-mail address with \email.
+%
+\alignauthor Junsung Kim\\
+       \affaddr{Dept. of Electrical and Computer Engineering}\\
+       \affaddr{Carnegie Mellon University}\\
+       \email{junsungk@cmu.edu}
+\alignauthor Ragunathan (Raj) Rajkumar\\
+       \affaddr{Dept. of Electrical and Computer Engineering}\\
+       \affaddr{Carnegie Mellon University}\\
+       \email{raj@ece.cmu.edu}
 \and
-Shinpei Kato\\
-\\
-Department of Computer Science\\
-University of California, Santa Cruz
+\alignauthor Shinpei Kato\\
+       \affaddr{Dept. of Information Engineering}\\
+       \affaddr{Nagoya University}\\
+       \email{shinpei@is.nagoya-u.ac.jp}
 }
+% There's nothing stopping you putting the seventh, eighth, etc.
+% author on the opening page (as the 'third row') but we ask,
+% for aesthetic reasons that you place these 'additional authors'
+% in the \additional authors block, viz.
+%\additionalauthors{Additional authors: John Smith (The Th{\o}rv{\"a}ld Group,
+%email: {\texttt{jsmith@affiliation.org}}) and Julius P.~Kumquat
+%(The Kumquat Consortium, email: {\texttt{jpkumquat@consortium.net}}).}
+%\date{30 July 1999}
+% Just remember to make sure that the TOTAL number of authors
+% is the number that will appear on the first page PLUS the
+% number that will appear in the \additionalauthors section.
 
@@
  developers can explicitly adjust the number of GPU cores depending on their needs. 
- An implicit adjustment will be supported by a run-time framwork, which 
+ An implicit adjustment will be supported by a run-time framework, which 
  dynamically allocates the number of cores to tasks based on the total 
@@
 Automatic safety features require smart planning and intelligent
-processing of data obtained from many sensors equiped in the vehicle,
+processing of data obtained from many sensors equipped in the vehicle,
 including LIDAR (LIght Detection And Ranging), radar, camera, and
@@
 of GPU cores depending on their needs. %An implicit adjustment will provide 
-A run-time framwork will dynamically
+A run-time framework will dynamically
 allocate the number of cores to tasks based on current 
@@
 
-The rest of paper is organized as follows.
-Section~\ref{sec:system_model} describes how our proposed system is modeled.
-Section~\ref{sec:adaptivity_support} presents the methods for adaptively managing GPU resources, and we conclude our paper in Section~\ref{sec:summary}.
+The rest of paper is organized as
+follows. Section~\ref{sec:system_model} describes how our proposed
+system is modeled. Section~\ref{sec:adaptivity_support} presents the
+methods for managing GPU resources in an adaptive manner, and we conclude our paper in Section~\ref{sec:summary}.
 
@@
 
-In this section, we describe adaptivity support for GPU
-applications.
+In this section, we describe adaptive support for GPU applications.
 We particularly focus on solving resource allocation problems.
@@
 We can cap the maximum number of cores available for an individual task
-to prevent it from abusing GPU resources, but the adaptivity of computing
+to prevent it from abusing GPU resources, but the adaptive nature of computing
 depends on the programmer, and outside system control.
@@
 
+In future work, we plan to integrate our frameworks into our prototype
+fault-tolerant adaptive systems~\cite{Kim_RTSS12} for autonomous driving
+technology.
+This is highly a challenging project to coordinate high-performance and
+adaptive computing for embedded real-time applications.
+
 \bibliographystyle{plain}
